\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-10-19 Paper/response-to-review.md

# Summary of changes

Thanks to the reviewers' comments, and some recent changes to ggplot2, the plotROC package has been greatly improved. The interface is now much more user-friendly, requires less typing, and is completely compatible with ggplot2. The computation is more efficient, and the plotting tools are much more flexible. I believe the package is much closer to my goal of providing a better tool for plotting ROC curves. Here are the main changes:

- The calculate_roc, ggroc, and plot_journal_roc functions are deprecated. 
- I have implemented the geom_roc, geom_rocci geometric layers, and stat_roc, stat_rocci statistical transformations. A recent update to ggplot2 now exports the Geom and Stat functions so that developers can create their own and be compatible with the other features of ggplot2. This is what we have done instead of the previous approach of "hacking" the existing ggplot2 geoms. The usage is now simplified to:

```r
p1 <- ggplot(data, aes(m = M, d = D)) + geom_roc() + geom_rocci() 
```

@@

 - The stat functions perform the calculations and the geoms add the visual elements. The user only has to specify the marker value m and binary outcome d, optionally with grouping or faceting. 
 - The resulting ggplot object can be passed to the plot_interactive_roc or export_interactive_roc functions to create interactive figures. Grouping and faceting are fully supported. 
 - The geom_roc displays cutoff labels by default. We have included an additional function style_roc() that adds the layer of styling to match the previous plot_journal_roc(). Its usage is as simple as adding it to a ggplot object: p1 + style_roc(). 
 - Direct labels can be added now with the direct_label() function. This takes the ggplot object as an argument so that it can extract the names: direct_label(p1). 
 - The AUC can be calculated by using the calc_auc() function on a ggplot object: calc_auc(p1). It calculates the AUC for each group or facet present in the plot. 
 - We provide the convenience function melt_roc(), which takes a dataset with multiple markers in different columns, and converts it to long format for use with the ggplot function. An example of its usage is in the paper.

The new version is not currently on CRAN, but I expect it will be in the next few months after ggplot2 is updated. To install, follow these two steps:

```r
devtools::install_github("hadley/ggplot2")
devtools::install_github("sachsmc/plotROC")
```

@@

# Reviewer 1

> This is a really cool package that uses state of the art tech with ggplot and shiny, as well as interactive plots, to display ROC curves.
> Unfortunately, the package is rather difficult to use and quite inconsistent. It also produces some bogus results (see below). As it is now, it looks more like a pretty cool tech demo than a scientific software I would use in my own research. Here are a few comments to try to improve it, as well as the manuscript itself.

## Bug report:
> The confidence intervals are apparently bogus. The following code executed in Rstudio or from the command line, with R 3.1.3 on Ubuntu 14.04 (from CRAN's Debian repository) and R 3.1.2 on Mac OS Yosemite (from Fink):

```r
set.seed(42)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
roc.estimate <- calculate_roc(round(M.ex, 2), D.ex, ci = TRUE)
plot_interactive_roc(ggroc(roc.estimate, ci = TRUE))
```

@@

> After clicking on a point around 1.6, I got the screenshot shown in Figure 1. This seems related to the rounding (here to the 2nd decimal).

Thank you for the bug report. I believe this was due to the way the previous function handled ties in the results. The computation of the ROC curve in R and the javascript functions for interactivity have been completely revamped. I cannot reproduce the error using the new update in any browser that I've tested, so I believe it has been fixed.

## General:
> • The manuscript is written partially in the first person. It is surprising and not totally consistent throughout the manuscript, some of which is written in passive tense (for instance “Labels can be added easily [...]”).

I have edited the manuscript to be more consistent with the use of active voice.

@@

This stub has been updated with descriptions, links to key functions, and a series of examples.

> • It is of course a design choice, but the use of underscores over dots (or camelCase) in function names seems quite surprising and may disturb some users. See Bååth, R. (2012). The State of Naming Conventions in R. R Journal, 4, 74-75. R Foundation for Statistical Computing for a review of current usage of function name styles.

This was a personal choice and unfortunately it won't please everyone. I opted to be consistent with the ggplot2 package, which uses underscores for functions and dots for objects and argument names.

> • Wouldn't it be more idiomatic to introduce the roc as a ggplot geometry? That would allow some much more straightforward syntax like:
> ```r
> data.for.roc <- data.frame(M.ex, D.ex)
> qplot(M.ex, D.ex, data = data.for.roc) + geom_roc()
> ```
>or suchlike that would be much easier to remember.

@@

It was my intention to develop a useful tool for plotting, which seemed to be lacking from the other packages that perform many useful statistical features as you have mentioned. I have included an additional function for calculating the AUC, but I think the other features are best left to the other packages. With the recent update to ggplot2, it is now possible to add additional stats and geoms. A potential future line of work would be to implement some of the statistical transformations and alternative estimates for ROC curves.

> Also the paper and package suffer from a number of ambiguities and inaccuracies (either technical or in the description):
	- The extreme point TPF/TPR=1/1 is not included in the output of the calculate_roc function. This could lead to problems when plotting curves that don't span the full 0/0 to 1/1 space and also when computing AUCs.

This has been corrected in the updated version. The (0,0) and (1,1) points are now always included in the computation output.

>	- plotROC seems to implictly define that samples ">" the threshold are positive, samples "<=" the threshold are negative. In this it differs from e.g. Fawcett "An introduction to ROC analysis" or other software (e.g. the ROCR package) that define samples ">=" the threshold as a positive prediction.

@@

This was an oversight and I thank you for pointing out the error. This has been corrected in the computation and in the text of the introduction.

>	- It is not obvious how plotROC handles ties (i.e. samples with identical numerical measurements)

The computation handles ties correctly by including them in the calculation of the proportions, and then removing the duplicated elements from the results. This is the same as what is implemented in the ROCR package.

>	- The paper defines "good outcome" and "bad outcome" in Section 1.1. These definitions seem to ignore half of the confusion matrix (i.e. ignoring true and false negatives).

@@

This section has been revamped and edited for clarity. The good and bad outcome terminology has been removed and the confusion matrix illustrating the concepts has been added as table 1.

>	- In "Help and Documentation" of the webpage, the confusion matrix and the definitions seem to be copied from Wikipedia. Credits and reference are needed.

### 2015-10-19 Paper/sachs2015plotROC.Rmd
Revised and resubmitted

\begin{table}
\begin{tabular}{l|c|c|c}
\multicolumn{1}{c}{}&\multicolumn{2}{c}{$D$}&\\
\cline{2-3}
\multicolumn{1}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
\cline{2-3}
 $M \geq c$ & true positive & false positive & test positive\\
\cline{2-3}
 $M < c$ & false negative & true negative & test negative\\
\cline{2-3}
\multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{}\\
\end{tabular}
\caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
\end{table}

@@

\pkg{plotROC} can be installed from the Comprehensive \proglang{R} Archive Network, or installed from source.

```{r load, eval = FALSE}
install.packages("plotROC")
library(plotROC)
```

@@

```{r inter, eval = FALSE}
plot_interactive_roc(basicplot)
plot_interactive_roc(basicplot, hide.points = TRUE)
plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw()))
```

@@

An interactive ROC plot can be exported by using the \code{export_interactive_roc} function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}. The \proglang{JavaScript} source can be omitted by using the \code{omit.js = TRUE} option. You may wish to do this when there are multiple interactive figures in a single document; the source only needs to be included once. The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr]. In a \pkg{knitr} document, it is necessary to use the \code{cat} function on the results and use the chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so that the interactive plot is displayed correctly. For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the \code{prefix} argument of \code{export_interactive_roc}. This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements. For examples of interactive plots and how to incorporate them into \pkg{knitr} documents, see the package vignette (\code{vignette("examples", package = "plotROC")}) or the web page https://sachsmc.github.io/plotROC/. The next code block shows an example \pkg{knitr} chunk that can be used in an .Rmd document to display an interactive plot.

@@

If you have grouping factors in your dataset, or you have multiple markers measured on the same subjects, you may wish to plot multiple ROC curves on the same plot. \pkg{plotROC} fully supports faceting and grouping done by \pkg{ggplot2}. In out example dataset, we have 2 markers measured in a paired manner:

@@

These data are in wide format, with the 2 markers going across 2 columns. \pkg{ggplot2} requires long format, with the marker result in a single column, and a third variable identifying the marker. We provide the convenience function \code{melt_roc} to perform this transformation. The arguments are the data frame, a name or index identifying the disease status column, and a vector of names or indices identifying the the markers. Optionally, the names argument gives a vector of names to assign to the marker, replacing their column names. The result is a data frame in long format.

@@

Interactive versions of the plots with grouping and faceting are fully supported.

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be plotted. Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way. A figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}. \pkg{plotROC} provides the convenience function \code{calc_auc} that takes a ggplot object that has an Roc layer, extracts the data, and calculates the AUC.

>>>>

\begin{table}
\begin{center}
\begin{tabular}{l|c|c|c}
\multicolumn{1}{c}{}&\multicolumn{2}{c}{$D$}&\\
\cline{2-3}
\multicolumn{1}{c|}{}&1&0&\multicolumn{1}{c}{Total}\\
\cline{2-3}
 $M \geq c$ & true positive & false positive & test positive\\
\cline{2-3}
 $M < c$ & false negative & true negative & test negative\\
\cline{2-3}
\multicolumn{1}{c}{Total} & \multicolumn{1}{c}{disease positive} & \multicolumn{1}{c}{disease negative} & \multicolumn{1}{c}{}\\
\end{tabular}
\caption{\label{confus} Confusion matrix illustrating the possible classification outcomes of a continuous test $M$ with threshold $c$ versus the true outcome $D$. }
\end{center}
\end{table}

@@

Users can install the latest version of \pkg{plotROC} from github, as shown below. The package will appear on the Comprehensive \proglang{R} Archive Network in the near future. It can also be downloaded from https://github.com/sachsmc/plotROC and installed from source. As of October 19, 2015, \pkg{plotROC} also requires the development version of \pkg{ggplot2} also available from github.

```{r load, eval = FALSE}
devtools::install_github("hadley/ggplot2")
devtools::install_github("sachsmc/plotROC")
library(plotROC)
```

@@

```{r inter, eval = FALSE}
plot_interactive_roc(basicplot)
plot_interactive_roc(basicplot, hide.points = TRUE)
plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw))
```

@@

Users can export an interactive ROC plot by using the \code{export_interactive_roc} function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}. The \proglang{JavaScript} source can be omitted by using the \code{omit.js = TRUE} option. Users may wish to do this when there are multiple interactive figures in a single document; the source only needs to be included once. The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr]. In a \pkg{knitr} document, it is necessary to use the \code{cat} function on the results and use the chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so that the interactive plot is displayed correctly. For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the \code{prefix} argument of \code{export_interactive_roc}. This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements. For examples of interactive plots and how to incorporate them into \pkg{knitr} documents, see the package vignette (\code{vignette("examples", package = "plotROC")}) or the web page https://sachsmc.github.io/plotROC/. The next code block shows an example \pkg{knitr} chunk that can be used in an .Rmd document to display an interactive plot.

@@

If you have grouping factors in your dataset, or you have multiple markers measured on the same subjects, you may wish to plot multiple ROC curves on the same plot. \pkg{plotROC} fully supports faceting and grouping as done by \pkg{ggplot2}. In my example dataset, there are 2 markers measured in a paired manner:

@@

These data are in wide format, with the 2 markers going across 2 columns. \pkg{ggplot2} requires long format, with the marker result in a single column, and a third variable identifying the marker. I provide the convenience function \code{melt_roc} to perform this transformation. The arguments are the data frame, a name or index identifying the disease status column, and a vector of names or indices identifying the the markers. Optionally, the names argument gives a vector of names to assign to the marker, replacing their column names. The result is a data frame in long format.

@@

This example is displayed in Figure \ref{multi}. Interactive versions of the plots with grouping and faceting are fully supported.

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be plotted. Therefore, users can add themes and annotations in the usual \pkg{ggplot2} way. A figure with a new theme, title, axis label, and AUC annotation is shown in Figure \ref{annotate}. \pkg{plotROC} provides the convenience function \code{calc_auc} that takes a ggplot object that has an Roc layer, extracts the data, and calculates the AUC.

### 2015-10-19 Paper/sachs2015plotROC.tex
Revised and resubmitted

\pkg{plotROC} can be installed from the Comprehensive \proglang{R}
Archive Network, or installed from source.

\begin{Schunk}
\begin{Sinput}
R> install.packages("plotROC")
R> library(plotROC)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> plot_interactive_roc(basicplot)
R> plot_interactive_roc(basicplot, hide.points = TRUE)
R> plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw()))
\end{Sinput}
\end{Schunk}

@@

An interactive ROC plot can be exported by using the
\code{export_interactive_roc} function, which returns a character string
containing the necessary \proglang{HTML} and \proglang{JavaScript}. The
\proglang{JavaScript} source can be omitted by using the
\code{omit.js = TRUE} option. You may wish to do this when there are
multiple interactive figures in a single document; the source only needs
to be included once. The character string can be copy-pasted into an
html document, or better yet, incorporated directly into a dynamic
document using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it
is necessary to use the \code{cat} function on the results and use the
chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so
that the interactive plot is displayed correctly. For documents that
contain multiple interactive plots, it is necessary to assign each plot
a unique name using the \code{prefix} argument of
\code{export_interactive_roc}. This is necessary to ensure that the
\proglang{JavaScript} code manipulates the correct svg elements. For
examples of interactive plots and how to incorporate them into
\pkg{knitr} documents, see the package vignette
(\code{vignette("examples", package = "plotROC")}) or the web page
https://sachsmc.github.io/plotROC/. The next code block shows an example
\pkg{knitr} chunk that can be used in an .Rmd document to display an
interactive plot.

>>>>

Users can install the latest version of \pkg{plotROC} from github, as
shown below. The package will appear on the Comprehensive \proglang{R}
Archive Network in the near future. It can also be downloaded from
https://github.com/sachsmc/plotROC and installed from source. As of
October 19, 2015, \pkg{plotROC} also requires the development version of
\pkg{ggplot2} also available from github.

\begin{Schunk}
\begin{Sinput}
R> devtools::install_github("hadley/ggplot2")
R> devtools::install_github("sachsmc/plotROC")
R> library(plotROC)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> plot_interactive_roc(basicplot)
R> plot_interactive_roc(basicplot, hide.points = TRUE)
R> plot_interactive_roc(basicplot, style = style_roc(theme = theme_bw))
\end{Sinput}
\end{Schunk}

@@

Users can export an interactive ROC plot by using the
\code{export_interactive_roc} function, which returns a character string
containing the necessary \proglang{HTML} and \proglang{JavaScript}. The
\proglang{JavaScript} source can be omitted by using the
\code{omit.js = TRUE} option. Users may wish to do this when there are
multiple interactive figures in a single document; the source only needs
to be included once. The character string can be copy-pasted into an
html document, or better yet, incorporated directly into a dynamic
document using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it
is necessary to use the \code{cat} function on the results and use the
chunk options \code{results = 'asis'} and \code{fig.keep = 'none'} so
that the interactive plot is displayed correctly. For documents that
contain multiple interactive plots, it is necessary to assign each plot
a unique name using the \code{prefix} argument of
\code{export_interactive_roc}. This is necessary to ensure that the
\proglang{JavaScript} code manipulates the correct svg elements. For
examples of interactive plots and how to incorporate them into
\pkg{knitr} documents, see the package vignette
(\code{vignette("examples", package = "plotROC")}) or the web page
https://sachsmc.github.io/plotROC/. The next code block shows an example
\pkg{knitr} chunk that can be used in an .Rmd document to display an
interactive plot.
